---
title: "Data Wrangling with *dplyr* and *tidyr*"
author: "Narayanan Raghupathy"
date: "September 13, 2016"
output: html_document
---

This R Markdown document covers some of the basics of data wrangling with *dplyr* and *tidyr* using the Software Carpentry [lessons](http://swcarpentry.github.io/r-novice-gapminder/) using gapminder data.

### Getting the Packages Needed

Let us load (or install) dplyr and tidyr, the R pacakges needed for the session

```{r load_r_packages}
#install.packages("tidyr")
#install.packages("dplyr")
library("tidyr")
library("dplyr")
```

We will be using the [gapminder](https://www.gapminder.org/) data set. We will first use use the  

*  [gapminder-FiveYearData](https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv)


and then we will use the 

* [gapminder-FiveYearData-Wide](https://raw.githubusercontent.com/rachelss/r-novice-gapminder/gh-pages/data/gapminder_wide.csv)

[gapminder-FiveYearData](https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv) is stored as CSV file. Let us first load it using **read.csv** command 
 
```{r load_gapminder_data}
gapminder <- read.csv("gapminder-FiveYearData.csv",stringsAsFactors = FALSE)
```

and examine the structure of the data using **str* command. We can see that this data frame contains five columns, with names

* country
* year
* pop
* continent
* lifeExp
* gdpPercap

```{r explore_gapminder_data}
print(str(gapminder))
```

```{r head_gapminder_data}
print(head(gapminder))
#print(dim(gapminder))
typeof(gapminder$gdpPercap)
```

## What is the average GDP per capita per continent?

Let us say, we want to compute average GDP of continents using the dataframe.

### Computing mean GDP of Continents: Old way


```{r mean_gdp1}
head(gapminder)
head(gapminder[gapminder$continent=="Asia",])
head(gapminder[gapminder$continent=="Asia","gdpPercap"])
mean(gapminder[gapminder$continent=="Asia","gdpPercap"])
mean(gapminder[gapminder$continent=="Africa","gdpPercap"])
mean(gapminder[gapminder$continent=="Americas","gdpPercap"])
```

### Data Wrangling with *dplyr*

With dplyr package, we can manipulate dataframes much easily without repeating ourselves too much. dplyr packages has the following six basic functions/operations,

* select()
* filter()
* summarize()
* mutate()
* group_by()

We will use some of these functions to manipulate the data frame and get computing the mean GDP of continents with dplyr.

### Select specific columns of a dataframe

```{r dplyr_select}
year_country_gdp <- select(gapminder,year,country,gdpPercap) 
head(year_country_gdp)
```

### %>%: Pipe operator
#### Select specific columns of a dataframe using %>%, the pipe operator

```{r dplyr_select_with_pipe}
year_country_gdp <- gapminder %>% select(year,country,gdpPercap)
head(year_country_gdp )
dim(year_country_gdp )
```

### Using *filter*

```{r dplyr_dilter_select_pipe}
year_country_gdp_euro <- gapminder %>%  
    filter(continent=="Europe") %>%
    select(year,country,gdpPercap)
dim(year_country_gdp_euro )
head(year_country_gdp_euro)
```
### *dplyr* challenge: using *filter* and *select*

Use *dplyr*'s *filter* and *select* functions to get year-wise life expectancy for the countries in Africa.

```{r dplyr_pipe_challenge1}
year_country_life_africa <- gapminder %>%  
    filter(continent=="Africa") %>%
    select(year,country,lifeExp)
head(year_country_life_africa)
dim(year_country_life_africa)
```

### Using *group_by* function

```{r dplyr_group_by}
head(gapminder %>% group_by(continent))
```
### Compute mean GDP of all continents with group_by and summarize

```{r mean_gdp_continent_dplyr}
gapminder %>% group_by(continent) %>%    summarize(mean_gdpPercap=mean(gdpPercap))
```

```{r mean_lifeexp_by_country}
mean_lifeexp_country=gapminder %>% group_by(country) %>%  summarize(mean_lifeExp=mean(lifeExp))
```

```{r_dplyr_example_group_by_2}
gdp_pop_bycontinents_byyear <- gapminder %>% group_by(continent,year) %>%
    summarize(mean_gdpPercap=mean(gdpPercap),
              sd_gdpPercap=sd(gdpPercap),
              mean_pop=mean(pop),
              sd_pop=sd(pop))
print(head(gdp_pop_bycontinents_byyear))
gdp_pop_bycontinents_byyear <- gapminder %>% 
                  group_by(continent,year) %>%
                  summarize(mean_gdpPercap=mean(gdpPercap),
                            mean_pop=mean(pop))
head(gdp_pop_bycontinents_byyear)
```

```{r_example3}
gdp_pop_bycontinents_byyear <- gapminder %>%
    mutate(gdp_billion=gdpPercap*pop/10^9) %>%
    group_by(continent,year) %>%
    summarize(mean_gdpPercap=mean(gdpPercap),
              sd_gdpPercap=sd(gdpPercap),
              mean_pop=mean(pop),
              sd_pop=sd(pop),
              mean_gdp_billion=mean(gdp_billion),
              sd_gdp_billion=sd(gdp_billion))
print(gdp_pop_bycontinents_byyear)
```


```{r wide_data}
head(gapminder)
dim(gapminder)
gap_wide <- read.csv("gapminder_wide.csv",stringsAsFactors = FALSE)
head(gap_wide)
dim(gap_wide)
gap_long <- gap_wide %>%
    gather(obstype_year, obs_values, starts_with('pop'),
           starts_with('lifeExp'), starts_with('gdpPercap'))
head(gap_long)
dim(gap_long)
gap_long <- gap_wide %>% gather(obstype_year,obs_values,-continent,-country)
gap_long <- gap_long %>% separate(obstype_year,into=c('obs_type','year'),sep="_")
str(gap_long)
gap_long$year <- as.integer(gap_long$year)
gap_long %>% group_by(continent,obs_type)%>%  summarize(means=mean(obs_values))
head(gap_long %>% spread(obs_type,obs_values) %>% arrange(country,continent,year))
head(gap_long)
head(gap_long %>% spread(obs_type,obs_values) %>% arrange(country,continent,year))
```
